Data migration under a Changing Schema

### 2023-03-12 2022 Migration/articleMigration.tex
define constraints for the type checker

\subsection{Changing data}
\label{sct:Changing data}
   Events can cause data to change.
   For triples, we distinguish just two types of event: insert and delete events.
   An event is a label (ins or del) with a dataset that is to be inserted in or deleted from another dataset.
   So we shall write
\begin{eqnarray}
   \ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}\pair{\triples}{\inst}&=&\pair{\triples\cup\triples_{\id{ev}}}{\inst\cup\inst_{\id{ev}}}\\
   \del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}\pair{\triples}{\inst}&=&\pair{\triples-\triples_{\id{ev}}}{\inst-\inst_{\id{ev}}}
\end{eqnarray}
   Insert events are represented as $\ins{\dataset}$ and delete events by $\del{\dataset}$.
   Both are functions $\Dataset\rightarrow\Dataset$, which are defined as:
   Since datasets are sets, we will use set union ($\cup$) and set difference ($-$) to make changes in datasets.

\subsection{Schemas}
\label{sct:Schemas}
   To design an information system, a developer defines a schema in which she defines concepts, specializations, relations, and rules.
   Together, these items capture the semantics of an information system~\cite{Spivak2012}.
   Each rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
   Every rule is an element of an infinite set called $\Rules$,
   which is disjoint from $\Atoms$, $\Concepts$, $\Rels$, and $\Dataset$.
   In this paper, variables $u$ and $v$ represent rules.

   A type system in Ampersand~\cite{vdWoude2011} identifies certain rules that are provably false.
   It also ensures that all rules can be interpreted within the set of relations in the schema.
   Due to this type system, the Ampersand compiler only generates code for scripts without type errors.
   As a result, for every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that says whether dataset $\dataset$ satisfies rule $u$.

@@

   Rules serve as invariants,
   i.e.\ constraints that the data in the database must satisfy at all times.
   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.

   A schema $\schema$ is a tuple $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
   in which $\concepts$ is a finite set of concepts,
   $\isa$ is a partial order of concepts,
   $\rels$ is a finite set of relations,
   and $\rules$ is a finite set of rules.
   The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
   Disambiguation sometimes requires to write $\concepts_{\schema}$, $\isa_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
   rather than $\concepts$, $\isa$, $\rels$, and $\rules$ respectively.
%  Every schema is an element of an infinite set called $\Schema$.

   In order to do static type checking,
   all concepts and relations must be ``known'' in the schema:
\begin{eqnarray}
   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:isasIntroduceConcepts}\\
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
\end{eqnarray}
   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~\ref{eqn:relationsIntroduceConcepts},
   this triple is said to {\em befit} schema $\quadruple{\concepts}{\isa}{\rels}{\rules}$.

@@

   The system is {\em at rest} when every rule in the schema is satisfied:
\begin{equation}
   \forall u\in\rules:\ \sat{u}{\dataset}\label{eqn:satisfaction}
\end{equation}
   However, the system continually registers events that insert or delete triples in its dataset,
   which may cause rules to be broken.
   This requires a reaction to restore invariance.
   Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
   Sending a message that acknowledges this receipt (another event) will cause the rule to be satisfied again.
   During the time between the receipt of the request and sending the acknowledgment, the rule is (temporarily) not satisfied.

@@

>>>>

\subsection{Schemas}
\label{sct:Schemas}
   To design an information system, a developer defines a schema in which she defines concepts, specializations, relations, and rules.
   Together, these items capture the semantics of an information system~\cite{Spivak2012}.
   A schema $\schema$ is a tuple $\quadruple{\concepts}{\isa}{\rels}{\rules}$,
   in which $\concepts$ is a finite set of concepts,
   $\isa$ is a partial order of concepts,
   $\rels$ is a finite set of relations,
   and $\rules$ is a finite set of rules.
   Each rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
   Every rule is an element of an infinite set called $\Rules$,
   which is disjoint from $\Atoms$, $\Concepts$, $\Rels$, and $\Dataset$.
   In this paper, variables $u$ and $v$ represent rules.
   The notation $\schema_{\infsys}$ refers to the schema of information system $\infsys$.
   Disambiguation sometimes requires to write $\concepts_{\schema}$, $\isa_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$
   rather than $\concepts$, $\isa$, $\rels$, and $\rules$ respectively.
   For every rule $u$ in the schema, there is a predicate $\sat{u}{\dataset}$ that says whether or not dataset $\dataset$ satisfies rule $u$.
%  Every schema is an element of an infinite set called $\Schema$.

   A schema must ensure that triples ``make sense'' in the semantics defined by the schema.
   This means that it must satisfy the following requirements.
\begin{eqnarray}
   \declare{n}{A}{B}\in\rels&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   A\ \isa\ B&\Rightarrow&A\in\concepts\ \wedge\ B\in\concepts
   \label{eqn:isasIntroduceConcepts}\\
   \isa&&\text{is a partial order on $\concepts$}
   \label{eqn:isasPartialOrder}\\
   \forall u\in\rules,\exists\dataset\in\Dataset-\emptyset&:&\sat{u}{\dataset}
   \label{eqn:trueifyableRules}
\end{eqnarray}
   Note that these requirements do not depend on any particular dataset,
   so they can be checked statically.
   The type system of Ampersand~\cite{vdWoude2011} ensures that every information system it generates satisfies these requirements.

@@

@@

   The information system enforces the semantics by ensuring that the following requirements remain true at all times:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}\\
   \forall u\in\rules&:&\sat{u}{\dataset}
   \label{eqn:satisfaction}
\end{eqnarray}
   Requirement~\ref{eqn:define R} ensures that the information system only works with triples whose relation is defined in the schema.
   Requirement~\ref{eqn:satisfaction} ensures that the dataset satisfies all rules in the schema.

   The maintainance relation between roles and rules is used to determine who is allowed to change the dataset.
   For every rule $u$ in the schema, there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   that represents the set of violations%

   The rules in the schema serve as invariants,
   i.e.\ constraints that the data in the database must satisfy at all times.
   At times in which the dataset does not satisfy a rule, that rule is said to be {\em broken}.
   In those cases, something or someone must take action to fix the situation and restore invariance of that rule.

   If a triple $\triple{a}{r}{b}$ satisfies requirement~\ref{eqn:define R} and its relation $r$ satisfies requirement~\ref{eqn:relationsIntroduceConcepts},
   this triple is said to {\em befit} schema $\quadruple{\concepts}{\isa}{\rels}{\rules}$.

   The system is {\em at rest} when every rule in the schema is satisfied:
\begin{equation}
\end{equation}
   However, the system continually registers events that insert or delete triples in its dataset,
   which may cause rules to be broken.
   This requires a reaction to restore invariance.
   Example: if a rule says that every request must be acknowledged, the receipt of a request (which is an event) will violate this rule.
   Sending a message that acknowledges this receipt (another event) will cause the rule to be satisfied again.
   During the time between the receipt of the request and sending the acknowledgment, the rule is (temporarily) not satisfied.

@@

   \subsection{Changing data}
   \label{sct:Changing data}
   @Bas, In het onderstaande heb ik mij er nog wat te makkelijk vanaf gemaakt.
   Aandacht nodig.

      Events can cause data to change.
      So, we model events as functions $\Dataset\rightarrow\Dataset$.
      For triples, we distinguish just two types of event: insert and delete events.
      An event is a label (ins or del) with a dataset that is to be inserted in or deleted from another dataset.
      We denote events as $\ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}$ and $\del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}$.
      They are defined by:
   \begin{eqnarray}
      \ins{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}(\dataset)&=&\pair{\triples_{\dataset}\cup\triples_{\id{ev}}}{\inst_{\dataset}\cup\inst_{\id{ev}}}\\
      \del{\pair{\triples_{\id{ev}}}{\inst_{\id{ev}}}}(\dataset)&=&\pair{\triples_{\dataset}-\triples_{\id{ev}}}{\inst_{\dataset}-\inst_{\id{ev}}}
   \end{eqnarray}
      Insert events are represented as $\ins{\dataset}$ and delete events by $\del{\dataset}$.
      Both are functions $\Dataset\rightarrow\Dataset$, which are defined as:
      Since datasets are sets, we will use set union ($\cup$) and set difference ($-$) to make changes in datasets.
